---
title: "Final Paper"
author: "Noah Edwards-Thro"
date: "1/23/2022"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Schedule

1/24 - Methods of Sloan Paper 1/31 - New Methods 2/7 - Reproducability
Graphic 2/14 - Results Copied from Sloan Paper 2/21 - New Results 2/28 -
Clean Up Visuals 3/14 - Discussion Part 1 3/21 - Open Week 3/28 -
Introduction/Final Edits Part 1 4/4 - Final Edits Part 2

# 2. Methods

To recreate the Sloan model as close as possible, I endeavored to use as
close to the same data and methods that Kalman and Bosch used as I
could.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(scifigure)
exps <- init_experiments(nexp = 3, exp_names = c("Kalman and Bosch", "Replication", "Expansion"), stage_names = c("population", "question", "hypothesis", "data", "analyst", "code", "estimate", "claim"))
exps[c("population", "analyst", "code", "estimate"),2] <- "different"
exps[c("population","hypothesis","data","analyst","code", "estimate", "claim"),3] <- "different"
sci_figure(exps, stage_names = c("Population", "Question", "Hypothesis", "Variables", "Analyst", "Code", "Results", "Claim"))
```

## 2.1 Reproduction of Sloan Paper

### 2.1.1 Data

Kalman and Bosch manually scraped 10 years (ranging from the 2009-2018
seasons) of player statistics covering advanced aggregation statistics,
per possession statistics, and shot distribution statistics. Their data
consisted of 5512 observations with 73 variables where each observation
was a single season of a single player (if a player played all ten
seasons during this range, he would show up 10 times).

While I desired to stay as close to the process that Kalman and Bosch
used as possible, I had knowledge of a package in R called nbastatr that
pulled data directly from Sports Reference LLC (the same website that
Kalman and Bosch manually scraped their data from). In an effort to get
more experience working with this package, I decided to pull the data
from this package using the `bref_player_stats` function. Additionally,
the shot distribution statistics were unable to download via the
nbastatr package so I manually scraped them from Sports Reference LLC.
Finally, I used the `player_profiles` function in the nbastatr package
to download the heights (in inches) for all of the players.


My data consisted of `r nrow(exps)` observations and `r ncol(exps)` variables. Through some
exploration, I found that the discrepancy in observations has to do with
players who are traded. In my dataset, any player who is traded
mid-season will still only show up as one row, with that players' team
being "TOT" (signaling that the player played for multiple teams).
Looking at the data on the Sports Reference LLC website, it seems that
Kalman and Bosch likely had a separate row for each team that a player
played on in a single season (so if a player played games for three
separate teams, he would have three separte rows for that season in
Kalman and Bosch's data while he would only have 1 row in mine).

In Kalman and Bosch's analysis, they filtered the data so that only
observations with more than 30 games in a season are counted, ending
with 3,608 observations. After using the same filter, I ended with 3,676
observations, likely due to the difference in data format with traded
players (a traded player who plays 20 games for two different teams
doesn't show up in their analysis but will show up as playing 40 games
in mine).

### 2.1.2 Variable Selection

I used the same variables that Kalman and Bosch used in their model (with the exception of using per 36 minute statistics instead of per 100 possession statistics for points and field goal attempts). The
variables were a combination of offensive, defensive, and aggregate
statistics and all statistics were calculated as rates except for player
height. The drawback of these rate statistics is that they does not take
into account how much a player is on the court. For instance, a player
who plays 10 minutes a night could have the same rate statistics as a
player who plays 35 minutes a night, but we would consider these players
very different in their abilities (largely due to the 35 minute player
being able to play efficiently for 35 minutes).

+----------------+-------------------------------------------------+
| **Variable**   | **Description**                                 |
+================+=================================================+
| Height         | Player height, in inches                        |
+----------------+-------------------------------------------------+
| Offensive      | \% of available offensive rebounds a player     |
| Rebound Rate   | gets while on the floor                         |
+----------------+-------------------------------------------------+
| Defensive      | \% of available defensive rebounds a player     |
| Rebound Rate   | gets while on the floor                         |
+----------------+-------------------------------------------------+
| Assist Rate    | \% of teammate field goals that a player        |
|                | assisted while on the floor                     |
+----------------+-------------------------------------------------+
| Steal Rate     | \% of opponent possessions that end with a      |
|                | steal b the player while on the floor           |
+----------------+-------------------------------------------------+
| Block Rate     | \% of opponent field goal attempted blocked by  |
|                | the player while on the floor                   |
+----------------+-------------------------------------------------+
| Turnover Rate  | Turnovers committed per 100 offensive           |
|                | possessions                                     |
+----------------+-------------------------------------------------+
| Points         | Points scored per 100 offensive possessions     |
+----------------+-------------------------------------------------+
| Usage Rate     | \% of offensive team possessions used by the    |
|                | player while on the floor                       |
+----------------+-------------------------------------------------+
| Player         | Per-minute production standardized such that    |
| Efficiency     | the league average is 15                        |
| Rating         |                                                 |
+----------------+-------------------------------------------------+
| Free Throw     | Number of free throws made per field goals      |
| Rate           | attempted                                       |
+----------------+-------------------------------------------------+
| Free Throw     | Number of free throws made per free throw       |
| Percentage     | attempt                                         |
+----------------+-------------------------------------------------+
| Field Goals    | Number of field goals attempted per 100         |
| Attempted      | possessions                                     |
+----------------+-------------------------------------------------+
| 2FG%           | Number of two-point field goals made per        |
|                | attempt                                         |
+----------------+-------------------------------------------------+
| 3FG%           | Number of three-point field goals made per      |
|                | attempt                                         |
+----------------+-------------------------------------------------+
| 2FG Assist     | \% of two-point field goals that are assisted   |
| Rate           |                                                 |
+----------------+-------------------------------------------------+
| 3FGA%          | \% of field goal attempts that are three-point  |
|                | attempts                                        |
+----------------+-------------------------------------------------+
| Corner 3FGA%   | \% of three point-field goal attempts that are  |
|                | from the corner                                 |
+----------------+-------------------------------------------------+
| 3FG Assist     | \% of three-point field goals that are assisted |
| Rate           |                                                 |
+----------------+-------------------------------------------------+
| Dunk Attempt   | \% of all field goal attempts that are dunks    |
| Rate           |                                                 |
+----------------+-------------------------------------------------+
| 0-3 ft FGA%    | \% of all field goal attempts between zero and  |
|                | three feet from the basket                      |
+----------------+-------------------------------------------------+
| 3-10 ft FGA%   | \% of all field goal attempts between three and |
|                | ten feet from the basket                        |
+----------------+-------------------------------------------------+
| 10ft-3p FGA%   | \% of all field goal attempts between ten feet  |
|                | from the basket and the three-point line        |
+----------------+-------------------------------------------------+

: Recreation from Kalman and Bosch

### 2.1.3 Gaussian Mixture Clustering

After attempting K-means clustering and being unsatisfied with the
results, Kalman and Bosch pivoted to model-based clustering to cluster
the players in their dataset. Model based clustering, specifically
finite Gaussian mixture modeling, uses an expectation-maximization (EM)
algorithm to fit observations into clusters. An advantage of model-based
clustering is that it assigns "soft clusters", showing the probability
that each observation will be in each cluster. The figure below (INSERT
FIGURE), displays a graphical representation as to the clustering
distributions produced by Gaussian mixture modeling.

As done by Kalman and Bosch, I used the "mclust" package in R to
implement the Gaussian mixture clustering.

## 2.2 Expansion of Sloan Paper

### 2.2.1 Data Expansion

While Kalman and Bosch's original paper only used data from the
2009-2018 seasons, I wanted to see what predictions the model would make
on more recent seasons (2019-2021 seasons) given the great degree to
which the NBA has changed over the past half decade alone. As before, I
used the `nbastatr` package to download the majority of the data and I
still had to manually scrape the shooting statistics necessary.

My three new seasons of data consisted of 1600 observations and 136
variables. To keep in line in the original analysis, I filtered the data
to players with greater than 30 games, resulting in 1125 observations
remaining.

### 2.2.2 Variable Reduction

With the new seasons of data, I first wanted to see what the predictions
looked like from the previous model. I used the same 23 variables and
ran the initial model on the new dataset.

Following this, I decided to change the input variables to the model and
reduce the number of clusters that it output. I believed that for the
most part, steal and block rates were mostly random and not necessarily
indicative of position (or had overlapping value with other variables -
e.g. height and block rates both being a strong predictor for
traditional center). I also cut out the PER variable as I believed that
it overlapped with other variables. Furthermore, while Kalman and Bosch
used solely rate statistics in their analysis, I sought to investigate
how game time (in minutes per game - MPG) could influence the model and
player predictions. In addition to adding this MPG variable to the
model, I combined some of the shooting variables into single variables
(10-16 and 16-3P became Midrange) in an effort to reduce the number of
input variables.

### 2.2.3 Cluster Reduction

Kalman and Bosch never outlined in their paper why they chose 9 clusters
as the optimal number, but I hypothesized that a slight reduction in
clusters (to 6 or 7) would make more intuitive sense.
