---
title: "Thesis Work"
author: Noah Edwards-Thro
output: github_document

---

9/21/21 - Downloaded Basketball Reference Data and set up github

603 rows of traded players, 39 double traded, 1 triple traded
603+39+1
643 +4762

5512 in the original Sloan Article. there are 5405 in ours. In data that matters (players with more than 30 games), we have slightly 68 more observations, likely again due to the traded feature, but working the opposite direction now. Say a player plays 25 games with one team and 25 games with another. The Sloan article won't pick that player up because he didn't play more than 30 games for either team. We will pick that player up because it will show up as 50 games for us. 

Because of the data structure (our data having one line per season even if the player is traded vs. their data likely having multiple lines per season if the player is traded), our density plots will be different.

Update 10/6 below

See new Geom_density with scraped data

Used Mclust
Also useful mclustBIC and mclustBICupdate
mclustModel

Potentially looking into doing dimension reduction in the future?

11/9/21

Initial model and second model don't really match. Even my initial model doesn't really match the output from the Sloan paper. 
Heat map seemed to relatively work. I messed around a little bit with a k means model just for the heck of it. Things to look at for next week are more heat map stuff and also how certain we are each player is in a certain group? Is that how our analysis can be a little bit different? Sum across teams and that way players with multiple parts show up?

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(nbastatR)
library(mclust)
```

```{r, eval = FALSE}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072*2)

bref_stats <- bref_players_stats(
  seasons = 2009:2018,
  tables = c("totals", "per_game", "advanced", "per_minute", "per_poss")
  )

write_csv(bref_stats, file = "bref_stats.csv")

bref_stats_new <- bref_players_stats(
  seasons = 2019:2021,
  tables = c("totals", "per_game", "advanced", "per_minute", "per_poss")
  )

write_csv(bref_stats_new, file = "bref_stats_new.csv")
```

```{r, message = FALSE, warning = FALSE}
bref_stats <- read_csv("bref_stats.csv")
bref_stats_new <- read_csv("bref_stats_new.csv")
```

```{r}
columns<-list("c","c","i","c","i","i","d","d","c","d","d","d","d","d","d","d","d","d","d","d","d","d","d","d", "d", "d", "d", "d", "d")
Shooting_Scrape <- read_csv("Basketball Reference Scrape.csv", col_types = columns)
Shooting_Scrape_New <- read_csv("~/Desktop/Fall 2021/Honors Thesis Work/Basketball Reference Scrape2.csv", col_types = columns)
```

```{r}
Shooting_Scrape <- Shooting_Scrape %>%
  separate(Player, c("Player_Name", "Player_ID"), sep = "\\\\")
Shooting_Scrape$yearSeason <- as.double(Shooting_Scrape$Year %>%
                                          str_extract(pattern = "[^-]+$"))
Shooting_Scrape <- Shooting_Scrape %>%
  group_by(Player_Name, Age) %>%
  filter(G == max(G))

Shooting_Scrape_New <- Shooting_Scrape_New %>%
  separate(Player, c("Player_Name", "Player_ID"), sep = "\\\\")
Shooting_Scrape_New$yearSeason <- as.double(Shooting_Scrape_New$Year %>%
                                          str_extract(pattern = "[^-]+$"))
Shooting_Scrape_New <- Shooting_Scrape_New %>%
  group_by(Player_Name, Age) %>%
  filter(G == max(G))

Player_Profile <- read_csv("Player_Profile.csv")
```

```{r, eval = FALSE}
Shooting_Scrape_Total <- rbind(Shooting_Scrape, Shooting_Scrape_New)
bref_stats_total <- rbind(bref_stats, bref_stats_new)
player_IDS <- bref_stats_total%>%select(idPlayerNBA)%>%unique()
player_IDS <- player_IDS[1281:1509,]

for (i in 1:length(player_IDS$idPlayerNBA)){
  player = player_profiles(player_ids = player_IDS$idPlayerNBA[i])
  player = player%>%select(idPlayer, namePlayer, heightInches)
  Player_Profile <- rbind(Player_Profile, player)
  print(i)
}
Player_Profile <- Player_Profile%>%unique()
write_csv(Player_Profile, "Player_Profile.csv")
```


# Evaluate if we have the correct number of players)
```{r, eval = FALSE}
not_traded <- bref_stats %>%
  filter(slugTeamBREF != "TOT")
traded <- bref_stats %>%
  filter(slugTeamBREF == "TOT")
traded_twice <- traded %>%
  filter(nchar(slugTeamsBREF) >9)
traded_thrice <- traded_twice %>%
  filter(nchar(slugTeamsBREF) > 15)
Count_Games <- bref_stats %>%
  filter(countGames > 30)
```

### Density Chart
```{r, eval = FALSE}
plot1 <- ggplot(Shooting_Scrape, aes(x = G)) +
  geom_density()
plot2 <- ggplot(bref_stats, aes(x = countGames)) +
  geom_density()
plot1
```

```{r}
order<-c(115,54,55,57,58,59,60,47,61,50,53,9,73,101,106,107,100,111,108,109,96,97,98,99)
order2<-c(2,1,3,11,12,109)
```


```{r}
master<-inner_join(
  bref_stats, 
  Shooting_Scrape, 
  by = c("yearSeason" = "yearSeason", "slugPlayerBREF"= "Player_ID"))
master <- left_join(master, Player_Profile, by = c("idPlayerNBA" = "idPlayer"))
master<-master%>%
  mutate(heightInches = case_when(
    namePlayer.x == "Andres Nocioni" ~ 79,
    namePlayer.x == "Anthony Randolph" ~ 82,
    namePlayer.x == "David Andersen" ~ 83,
    namePlayer.x == "Dionte Christmas" ~ 77,
    namePlayer.x == "Frank Mason III" ~ 71,
    namePlayer.x == "Gustavo Ayon" ~ 82,
    namePlayer.x == "Jeff Adrien" ~ 79,
    namePlayer.x == "Joey Dorsey" ~ 80,
    namePlayer.x == "Rudy Fernandez" ~ 78,
    namePlayer.x == "Samardo Samuels" ~ 81,
    namePlayer.x == "TJ Leaf" ~ 82,
    namePlayer.x == "Tobias Harris" ~ 80,
    namePlayer.x == "Tyler Hansbrough" ~ 81,
    namePlayer.x == "Victor Claver" ~ 81,
    namePlayer.x == "Vitor Luiz Faverani" ~ 83,
    TRUE ~ heightInches
  ))
nums <- unlist(lapply(master, is.numeric))
master2 <- master[nums]
master2 <- master2%>%
  filter(countGames >= 30)
master3 <- master2[order]
master5 <- master3 %>%
  mutate(across(everything(), .fns = ~replace_na(., 0)))
master6 <- master%>%
  filter(countGames >= 30)%>%
  select(all_of(order2))
```

```{r}
new_master<-inner_join(
  bref_stats_new, 
  Shooting_Scrape_New, 
  by = c("yearSeason" = "yearSeason", "slugPlayerBREF"= "Player_ID")
)
new_master <- left_join(new_master, Player_Profile, by = c('idPlayerNBA' = "idPlayer"))
new_master<-new_master%>%
  mutate(heightInches = case_when(
    namePlayer.x == "Frank Mason III" ~ 71,
    namePlayer.x == "Landry Shamet" ~ 82,
    namePlayer.x == "Nicolas Claxton" ~ 79,
    namePlayer.x == "PJ Dozier" ~ 80,
    namePlayer.x == "TJ Leaf" ~ 82,
    namePlayer.x == "Tobias Harris" ~ 80,
    TRUE ~ heightInches
  ))
new_master2 <- new_master%>%
  filter(countGames >= 30)
new_master3 <- new_master2[nums]
new_master4 <- new_master3[order]
new_master5 <- new_master4 %>%
  mutate(across(everything(), .fns = ~replace_na(., 0)))
new_master5 <- rbind(master5, new_master5)
new_master6 <- new_master2[order2]
new_master6 <- rbind(master6, new_master6)
```

```{r, eval = FALSE}
library(visdat)
vis_miss(new_master5, cluster = TRUE)
```


```{r mclust, cache = TRUE}
set.seed(21)
start_time <- Sys.time()
m <- Mclust(master5, G = 9)
end_time <- Sys.time()
end_time - start_time
```


```{r}
master7 <- master5
master7$first_pred <- m$classification
master7 <- cbind(master6, master7)
d <- purrr::map(1:9, ~ master7%>%select_if(is.numeric) %>% filter(first_pred == {{.x}}))
```

```{r mclust2, cache = TRUE}
set.seed(23)
start_time <- Sys.time()
m2 <- Mclust(new_master5, G = 9)
end_time <- Sys.time()
end_time - start_time
```

```{r}
new_predictions <- predict(m, newdata = new_master5[3711:4855,])
new_master7 <- new_master5
first_pred <- c(m$classification, new_predictions$classification)
new_master7$first_pred <- first_pred
new_master7$second_pred <- m2$classification
new_master7 <- cbind(new_master6, new_master7)
d2 <- purrr::map(1:9, ~ new_master7%>%select_if(is.numeric) %>% filter(second_pred == {{.x}}))
prop.table(table(new_master7$first_pred))

new_master7 <- new_master7%>%
  mutate(First_Prediction = case_when(
    first_pred == 1 ~ "Floor General",
    first_pred == 2 ~ "Stretch Forward",
    first_pred == 3 ~ "Mid Range Big",
    first_pred == 4 ~ "Traditional Center",
    first_pred == 5 ~ "Ball Dominant Scorer",
    first_pred == 6 ~ "Versatile Role Player",
    first_pred == 7 ~ "High Usage Guard",
    first_pred == 8 ~ "Skilled Forward",
    first_pred == 9 ~ "Three Point Shooting Guard"))
new_master7 <- new_master7%>%
  mutate(Second_Prediction = case_when(
    second_pred == 1 ~ "Mid Range Big",
    second_pred == 2 ~ "Stretch Forward",
    second_pred == 3 ~ "Skilled Forward",
    second_pred == 4 ~ "High Usage Guard",
    second_pred == 5 ~ "Ball Dominant Scorer",
    second_pred == 6 ~ "Traditional Center",
    second_pred == 7 ~ "Versatile Role Player",
    second_pred == 8 ~ "Floor General",
    second_pred == 9 ~ "Three Point Shooting Guard"))
new_master7 <- new_master7%>%
  mutate(Match = (First_Prediction == Second_Prediction))
mean(new_master7$Match)

new_master7%>%
  group_by(First_Prediction)%>%
  summarize(n = n())
new_master7%>%
  group_by(Second_Prediction)%>%
  summarize(n = n())
new_master7%>%
  filter(First_Prediction == "Ball Dominant Scorer" &
        Second_Prediction != "Ball Dominant Scorer")
```

```{r}
set.seed(123)
wss <- function(k) {
  kmeans(new_master5, k, nstart = 10 )$tot.withinss
}
k.values <- 1:10
wss_values <- map_dbl(k.values, wss)
plot(k.values, wss_values,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")
```



```{r}
pred_1_median<-map_df(d, ~ .x %>% summarise(across(everything(), median)))
pred_1_mean<-map_df(d, ~ .x %>% summarise(across(everything(), mean)))
pred_2_median<-map_df(d2, ~ .x %>% summarise(across(everything(), median)))
pred_2_mean<-map_df(d2, ~ .x %>% summarise(across(everything(), mean)))
```

```{r, eval = FALSE}
prop.table(table(m$classification))
prop.table(table(m2$classification))
```

```{r}
Teams <- new_master7%>%
  group_by(Tm, slugSeason, first_pred)%>%
  summarise(total = n())
  
ggplot(Teams%>%filter(Tm == "CHI"), aes(x = slugSeason, y = first_pred, fill = total)) +
  geom_tile()+
  scale_fill_gradient(low="pink", high="red")+
  scale_y_discrete(limits = c("1","2", "3", "4", "5", "6", "7", "8", "9"))+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, vjust = .5))
```

